On the error performance of the $A_n$ lattices

### 2012-03-08 paper.tex
more additions

This paper is organised as follows.  In Section~\ref{sec:latt-latt-codes} we give a brief overview of lattices and codes constructed from them, i.e., \emph{lattice codes}.  We describe lattice decoding and show how the probability of coding error can be expressed in terms of the moments of the \emph{Voronoi cell} of the lattice used.  Section~\ref{sec:main-result} states the main result, describing recursive formula to compute the moments of the Voronoi cell of $A_n$.  Section~\ref{sec:lattice-a_n} describes the lattice $A_n$ and some of its properties.  An important property for our purposes is that the Voronoi cell of $A_n$ is precisely the orthogonal projection of an $(n+1)$-dimensional hypercube onto a hyperplace orthogonal to one of its vertices~\cite{McKilliam2009CoxeterLattices,McKilliam2010thesis}.
 %This result was previously known~\cite{McKilliam2009CoxeterLattices,McKilliam2010thesis}.  The proof is short and we give it again here so that this paper is self contained.
In Section~\ref{sec:integr-funct-over} we use this property to show how integrals over the Voronoi cell of $A_n$ can be expressed as integrals over the $(n+1)$-dimensional hypercube. These integrals are solvable and we use them to obtain the moments of the Voronoi cell in Section~\ref{sec:powers-eucl-norm}.  In Section~\ref{sec:results-simulations} we plot the probability of error versus signal to noise ratio for codes constructed from the lattices $A_1 \simeq \ints$, $A_2$, $A_3 \simeq D_3$ and $A_4$, $A_5$ and $A_8$.  We also plot the results of Monte-Carlo simulations that support our analytical results.
%An advantage of the analytical results is that the probability of error can calculated at high SNR, when the probabilty of error is small.  This is in contrast to Monte-Carlo simulations that are computationally infeasible at when the probability of error is small.  We display the performance of these lattice codes

@@

%For each lattice point $\ybf \in \Lambda$ there exists a Voronoi cell containing those point from $\reals^n$ nearer to $\ybf$ that any other lattice point.  Due to the structure of the lattice the Voronoi cell of each lattice point is the same shape.  Denote by $\vor(\Lambda)$ the Vornoi cell of the lattice point at the origin. NEAREST POINT PROBLEM.
\newcommand{\calX}{\mathcal X}
\newcommand{\calC}{\mathcal C}
Lattices can be used to construct \emph{signal space codes}.  A signal space code $\calC$ of dimension $n$ is a finite set of vectors (or points) in the Euclidean space $\reals^n$.  Each vector in $\calC$ is called a \emph{codeword} and represents a particular signal.  The number of codewords is denoted by $\abs{\calC}$.  If each codeword is transmitted with equal probability then the rate of the code is $R = \frac{1}{n}\log_2\abs{\calC}$ bits per codeword. The average power of the code is given by $P = \frac{1}{\abs{\calC}}\sum_{\cbf \in \calC}\|\cbf\|^2$.  In the AWGN channel the received signal takes the form
\[
\ybf = \cbf + \wbf
\]
where $\ybf \in \reals^n$, $\cbf \in \calC$ and $\wbf$ is a vector of independent and identically distributed Gaussian random variables with variance $\sigma^2$.  If the receiver employs maximum likelihood decoding then the estimator of $\cbf$ given $\ybf$ at the receiver is
\begin{equation}\label{eq:mldecoder}
\hat{\cbf} = \underset{\cbf \in \calC}{\operatorname{argmin}}\;\| \ybf - \cbf \|^2.
%\hat{\cbf} &=& \mbox{arg~}\min_{\cbf \in \calC} \| \ybf - \cbf \|^2.
\end{equation}
That is, the receiver computes the codeword in $\calC$ nearest in Euclidean distance to the received signal $\ybf$.  Assuming that each codeword is transmitted with equal probability then the probability of correct decoding is
\[
P_C =\frac{1}{\abs{\calC}}\sum_{\cbf \in \calC} \prob( \hat{\cbf} = \cbf ),
\]
and the probability of error is
\[
P_E = 1 - \frac{1}{\abs{\calC}} \sum_{\cbf \in \calC} \prob(
\hat{\cbf} = \cbf ).
\]
%The maximum possible rate of reliable communications in this channel is given by the channel capacity $B\log_2\left(1 + \frac{P}{\sigma^2}\right)$.  If the codewords are chosen uniformly randomly on a sphere of radius $\sqrt{P}$ then as $n$ gets large the rate of the code can be made arbitrarily close to the channel capacity while simultaneously the probability of error can be made arbitrarily small.  Another way of stating this is that we can choose close to $2^{nR/B}$ codewords uniformly randomly on a sphere of radius $\sqrt{P}$ and as $n$ gets large the probability of error decreases to zero.

%A problem with randomly generated codes and maximum likelihood decoding is that there are no computationally efficient methods for encoding and decoding.  For encoding the transmitter must store a list of size $\abs{\calC} \approx 2^{nR/B}$ and for decoding the receiver must compute $\hat{\cbf}$ requiring a search over all $\abs{\calC} \approx 2^{nR/B}$ codewords.  As $n$ gets large this becomes infeasible.  To avoid this computational complexity it is preferable for a code to have some type of `structure' that is exploitable for efficient encoding and decoding.  One way of enforcing structure is to choose codewords from a lattice.  The idea being that encoding and decoding can now be performed efficiently by \emph{lattice decoding}, i.e. by finding a nearest point in a lattice.

A \emph{lattice code} is a signal space code with codewords taken from a finite subset of points of some lattice $\Lambda$ in $\reals^n$. There are infinitely many ways to choose a finite subset from a lattice, but common approaches make use of a bounded subset of $\reals^n$, called a \emph{shaping region} $S \subset \reals^n$.  The codewords are given by those lattice points inside the shaping region, that is,
\[
\calC = S \cap \Lambda.
\]
Common choices of shaping region are $n$-dimensional spheres, spherical shells, hypercubes, or the Voronoi cell of a \emph{sublattice} of $\Lambda$~\cite{Buda1989_some_opt_codes_structure,Erex2004_lattice_decoding,Conway1983VoronoiCodes}.  A consequence of the finiteness of a lattice code is that the maximum likelihood decoder~\eqref{eq:mldecoder} is \emph{not} equivalent to lattice decoding.  Computing a nearest lattice point will not in general return a lattice point from the code (the decoded lattice point might lie outside the shaping region).
 %A question that remained open for some time was whether lattice codes employing lattice decoding could achieve channel capacity.  This question was answered in the affirmative by Erez and Zamir~\cite{Erex2004_lattice_decoding}.

Let $\calC$ be a lattice code constructed from a lattice $\Lambda$.  Assuming that each codeword $\cbf \in \calC$ has an equal probability of being transmitted and that the receiver employs lattice decoding, then the average probability of correct decoding in the AWGN channel is
%\footnote{In order to approach capacity Erez and Zamir~\cite{Erex2004_lattice_decoding} employ a uniform dither and also a minimum mean square error scaling parameter $\alpha$ (the so called \emph{modulo-lattice additive noise channel}).  A result of this is that that observed noise distribution is no longer Gaussian. However, for large signal to noise ratio $\alpha \approx 1$ and the observed noise is very close to Gaussian.  In this case~\eqref{eq:PEgaussian} is a reasonable approximation.}
\begin{align}
%P_C &\geq P_C(\Lambda) = \lim_{\abs{\calC}\rightarrow \infty} \frac{1}{\abs{\calC}} \sum_{\cbf\in \calC} \prob( \cbf + \wbf \in \vor(\Lambda) + \cbf )  \nonumber \\
P_C &\geq P_C(\Lambda) = \frac{1}{\abs{\calC}} \sum_{\cbf\in \calC} \prob( \cbf + \wbf \in \vor(\Lambda) + \cbf )  \nonumber \\
&= \prob(\wbf \in \vor(\Lambda) ) \nonumber \\
&=   \frac{1}{(\sqrt{2\pi}\sigma)^n} \int_{\vor(\Lambda)}
e^{-\|\xbf\|^2 / 2\sigma^2 } d\xbf. \label{eq:PEgaussian}
\end{align}
The probability of error is $P_E \leq 1 - P_C(\Lambda)$, where the upper bound is asymptotically tight for large constellations (the proportion of the codewords near the boundary of the shaping region becomes small).  By expanding $e^x = 1  + x + \frac{x^2}{2} + \dots$ according to its Maclaurin series we obtain
\begin{align}
P_C(\Lambda)  &= \frac{1}{(\sqrt{2\pi}\sigma)^n}\int_{\vor(\Lambda)} 1
- \frac{\|\xbf\|^2}{2\sigma^2} + \frac{\left(\|\xbf\|^2\right)^2}{
4\sigma^42!} - \dots d\xbf \nonumber \\
&= \frac{1}{(\sqrt{2\pi}\sigma)^n} \sum_{m=0}^\infty
\frac{(-1)^m}{2^m\sigma^{2m}m!} \int_{\vor(\Lambda)} \|\xbf\|^{2m}
d\xbf.  \label{eq:summomentproberror}
\end{align}
So, to obtain arbitrarily accurate approximations to the probability of error it is enough to know the values of $\int_{\vor(\Lambda)} \|\xbf\|^{2m} d\xbf$ for $m=1,2\dots$ for some sufficiently large $m$.  The number of terms required will increase as the noise variance gets smaller.  This implies that the bound with a fixed number of terms is not asymptotically tight but is very accurate up to a finite signal-to-noise ratio.

\newcommand{\calM}{M}
In this paper we focus on $n$-dimensional lattice codes constructed from the family of lattices called $A_n$ and  we will derive expressions for the \emph{moment}s
\[
 \calM_n(m) = \int_{\vor(A_n)} \|\xbf\|^{2m} d\xbf.
\]
These can be summed in (\ref{eq:summomentproberror}) to give arbitrarily accurate approximations for the probability of error.

@@

\begin{figure*}[tp]
	\centering
		\includegraphics[scale=1.6]{figs/cuberothex-1.mps}
		\caption{Orthogonal projection of a cube as it is rotated about its center.  The figure of the left is the view of a cube from side on, and the boundary is a square.  When the cube is rotated the boundary becomes a hexagon.  The hexagon is regular when the cube is viewed along one of its vertices (the rightmost figure).  The regular hexagon is the Voronoi cell of the hexagonal lattice $A_2$.  Every 2-dimensional zonotope is also unimodular so the boundary of each of the figures above can be used to tile Euclidean space~\cite{McMullen_space_tiling_zonotopes_1975}.}
		\label{fig:cuberothex}
\end{figure*}

>>>>

This paper is organised as follows.  In Section~\ref{sec:latt-latt-codes} we give a brief overview of lattices and codes constructed from them, i.e., \emph{lattice codes}.  We describe lattice decoding and show how the probability of coding error can be expressed in terms of the moments of the \emph{Voronoi cell} of the lattice used.  Section~\ref{sec:main-result} states the main result, describing recursive formula to compute the moments of the Voronoi cell of $A_n$.  Section~\ref{sec:lattice-a_n} describes the lattice $A_n$ and some of its properties.  An important property for our purposes is that the Voronoi cell of $A_n$ is precisely the orthogonal projection of an $(n+1)$-dimensional hypercube onto a hyperplane orthogonal to one of its vertices~\cite{McKilliam2009CoxeterLattices,McKilliam2010thesis}.
 %This result was previously known~\cite{McKilliam2009CoxeterLattices,McKilliam2010thesis}.  The proof is short and we give it again here so that this paper is self contained.
In Section~\ref{sec:integr-funct-over} we use this property to show how integrals over the Voronoi cell of $A_n$ can be expressed as integrals over the $(n+1)$-dimensional hypercube. These integrals are solvable and we use them to obtain the moments of the Voronoi cell in Section~\ref{sec:powers-eucl-norm}.  In Section~\ref{sec:results-simulations} we plot the probability of error versus signal to noise ratio for codes constructed from the lattices $A_1 \simeq \ints$, $A_2$, $A_3 \simeq D_3$ and $A_4$, $A_5$ and $A_8$.  We also plot the results of Monte-Carlo simulations that support our analytical results.
%An advantage of the analytical results is that the probability of error can calculated at high SNR, when the probabilty of error is small.  This is in contrast to Monte-Carlo simulations that are computationally infeasible at when the probability of error is small.  We display the performance of these lattice codes

@@

%For each lattice point $\ybf \in \Lambda$ there exists a Voronoi cell containing those point from $\reals^n$ nearer to $\ybf$ that any other lattice point.  Due to the structure of the lattice the Voronoi cell of each lattice point is the same shape.  Denote by $\vor(\Lambda)$ the Vornoi cell of the lattice point at the origin. NEAREST POINT PROBLEM.
\newcommand{\calX}{\mathcal X}
\newcommand{\calC}{\mathcal C}
Lattices can be used to construct \emph{lattice codes}.  A lattice code $\calC$ of dimension $n$ is a finite subset of points of some lattice $\Lambda$ in $\reals^n$.  Each point in $\calC$ is called a \emph{codeword} and represents a particular signal.  There are infinitely many ways to choose a finite subset from a lattice, but common approaches make use of a bounded subset of $\reals^n$, called a \emph{shaping region} $S \subset \reals^n$.  The codewords are given by those lattice points inside the shaping region, that is, $\calC = S \cap \Lambda.$  Common choices of shaping region are $n$-dimensional spheres, spherical shells, hypercubes, or the Voronoi cell of a \emph{sublattice} of $\Lambda$~\cite{Buda1989_some_opt_codes_structure,Erex2004_lattice_decoding,Conway1983VoronoiCodes}.  The number of codewords is denoted by $\abs{\calC}$.  If each codeword is transmitted with equal probability then the rate of the code is $R = \frac{1}{n}\log_2{\abs{\calC}}$ bits per codeword.  The average power of the code is $P = \frac{1}{\abs{\calC}}\sum_{\cbf \in \calC}\|\cbf\|^2$.

In the AWGN channel the received signal takes the form
\[
\ybf = \cbf + \wbf
\]
where $\ybf \in \reals^n$, $\cbf \in \calC$ and $\wbf$ is a vector of independent and identically distributed Gaussian random variables with variance $\sigma^2$.  If the receiver employs maximum likelihood decoding then the estimator of $\cbf$ given $\ybf$ at the receiver is
\begin{equation}\label{eq:mldecoder}
\hat{\cbf}_{\text{ML}} = \underset{\cbf \in C}{\operatorname{argmin}}\;\| \ybf - \cbf \|^2,
%\hat{\cbf} &=& \mbox{arg~}\min_{\cbf \in \calC} \| \ybf - \cbf \|^2.
\end{equation}
that is, the receiver computes the codeword in $\calC$ nearest in Euclidean distance to the received signal $\ybf$.  Assuming that each codeword is transmitted with equal probability, then probability of correct maximum likelihood decoding is
\[
\sum_{\cbf \in \calC}\prob( \hat{\cbf}_{\text{ML}} = \cbf ).
\]
Typically maximum likelihood decoding is computationally complex and it is preferable to use lattice decoding~\cite{Agrell2002,Erex2004_lattice_decoding}.  The estimator of $\cbf$ is then,
\begin{equation}\label{eq:latticedecoder}
\hat{\cbf} = \underset{\cbf \in \Lambda}{\operatorname{argmin}}\;\| \ybf - \cbf \|^2,
%\hat{\cbf} &=& \mbox{arg~}\min_{\cbf \in \calC} \| \ybf - \cbf \|^2.
\end{equation}
that is, the receiver computes the lattice point in $\Lambda$ nearest in Euclidean distance to the received signal $\ybf$.  Equivalently, $\hat{\cbf}$ is the lattice point such that $\ybf \in \vor(\Lambda) + \hat{\cbf}$.  Note that with lattice decoding the decoded lattice point $\hat{\cbf}$ is not guaranteed to be inside the code $\calC$.

Correct lattice decoding occurs when $\hat{\cbf} = \cbf$, or equivalently when $\ybf \in \vor(\Lambda) + \cbf$, or equivalently when $\wbf \in \vor(\Lambda)$, i.e. when the noise $\wbf$ is inside the Voronoi cell of the lattice.  Assuming that each codeword is transmitted with equal probability then the probability of correct lattice decoding is
\begin{align}
P_C &= \sum_{\cbf \in \calC} \prob( \hat{\cbf} = \cbf ) \\
&= \sum_{\cbf\in \calC} \prob( \cbf + \wbf \in \vor(\Lambda) + \cbf )  \nonumber \\
&= \prob(\wbf \in \vor(\Lambda) ) \nonumber \\
&=   \frac{1}{(\sqrt{2\pi}\sigma)^n} \int_{\vor(\Lambda)}e^{-\|\xbf\|^2 / 2\sigma^2 } d\xbf. \label{eq:PEgaussian}
\end{align}
The probability of error is $P_E = 1 - P_C$.  The probability of correct lattice decoding is smaller than the probability of correct maximum likelihood decoding, however, as the size of the code $\abs{\calC}$ increases, $P_C$ converges to the probability of correct maximum likelihood decoding (the proportion of codewords near the boundary of the shaping region become small).

By expanding $e^x = 1  + x + \frac{x^2}{2} + \dots$ according to its Maclaurin series we obtain
\begin{align}
P_C  &= \frac{1}{(\sqrt{2\pi}\sigma)^n}\int_{\vor(\Lambda)} 1
- \frac{\|\xbf\|^2}{2\sigma^2} + \frac{\left(\|\xbf\|^2\right)^2}{
4\sigma^42!} - \dots d\xbf \nonumber \\
&= \frac{1}{(\sqrt{2\pi}\sigma)^n} \sum_{m=0}^\infty
\frac{(-1)^m}{2^m\sigma^{2m}m!} \int_{\vor(\Lambda)} \|\xbf\|^{2m}
d\xbf.  \label{eq:summomentproberror}
\end{align}
So, to obtain arbitrarily accurate approximations to the probability of error it is enough to know the values of $\int_{\vor(\Lambda)} \|\xbf\|^{2m} d\xbf$ for $m=1,2\dots$ for some sufficiently large $m$.  The number of terms required increases as the noise variance gets smaller.  This implies that the bound with a fixed number of terms is not asymptotically tight but is very accurate up to a finite signal-to-noise ratio.  We call these the \emph{moments} of $\vor(\Lambda)$.

\newcommand{\calM}{M}
In this paper we focus on $n$-dimensional lattice codes constructed from the family of lattices called $A_n$ and we derive expressions for the moments
\[
 \calM_n(m) = \int_{\vor(A_n)} \|\xbf\|^{2m} d\xbf.
\]
These can be summed in (\ref{eq:summomentproberror}) to give arbitrarily accurate approximations for the probability of error.

@@

\begin{figure*}[tp]
	\centering
		\includegraphics[scale=1.6]{figs/cuberothex-1.mps}
		\caption{Orthogonal projection of a cube as it is rotated about its center.  The figure of the left is the view of a cube from side on, and the boundary is a square.  When the cube is rotated the boundary becomes a hexagon.  The hexagon is regular when the cube is viewed along one of its vertices (the rightmost figure).  The regular hexagon is the Voronoi cell of the hexagonal lattice $A_2$.  Every 2-dimensional zonotope is also unimodular so the boundary of each of the figures above can be used to tile 2-dimensional Euclidean space~\cite{McMullen_space_tiling_zonotopes_1975}.}
		\label{fig:cuberothex}
\end{figure*}
